Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-06-17 paper.tex
fill in neand ancestry details

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150ka was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140ka. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), contemporary populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between contemporary populations can
be attributed to drift between stems (Figures~\ref{fig:predictions} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

>>>>

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150ka was remarkably robust.
In a reticulated model, we use ``divergence'' between populations to describe
the time of their most recent shared ancestry.
The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140ka. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), contemporary populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between contemporary populations can
be attributed to drift between stems (Figures~\ref{fig:predictions} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

Other studies have fit tree-like demographic models to African populations using 
distributions of allele frequencies or related statistics,
finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.
Section~\ref{sec:IM-reinfer} explains this behaviour:
split times inferred using isolation with migration models and
distributions of allele frequencies simulated under our reticulated
models are highly variable and much deeper than the more recent
estimated population divergence.

### 2022-06-17 supplement.tex
fill in neand ancestry details

>>>>

\subsection{Back-to-Africa Gene flow and Neanderthal ancestry in Africa}

Back-to-Africa migrations likely introduced Neanderthal ancestry into Africa
\citep{Chen2020-wh}. This Neanderthal ancestry would cause reticulation in the
ancestry of the African populations, and is therefore relevant to the modeling
presented here. The models we inferred account for both Neanderthal admixture
in Eurasia and Back-to-Africa migrations, such that they predict the amount of
Neanderthal ancestry that we would expect to find in African individuals. We
find that African populations without high proportions of recent Eurasian
ancestry are predicted to carry $0.1-0.2\%$ Neanderthal
ancestry (Table~\ref{tab:supp-lineage-probabilitys}),
which is remarkably consistent with the results of
\citet{Chen2020-wh}. Since the IBD statistics from \citet{Chen2020-wh} are not
used in the present analysis, this serves as an independent validation of the
present models. More importantly, since the Neanderthal admixture in Africa is
properly accounted for in the model, we do not expect the presence of such
ancestry to bias our estimate of reticulation in Africa.
